$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-07-10 paper/paper.tex
Introduce the de Bruijn indices to names and back subsection

% Constructors
\newcommand{\PO}{\constr{\mathbb{0}}}
\newcommand{\comp}[2]{#1 \, \constr{\parallel} \, #2}
\newcommand{\new}{\constr{\boldsymbol{\nu}} \,}
\newcommand{\send}[2]{#1 \, \constr{\langle} \, #2 \,\constr{\rangle} \,}
\newcommand{\sendp}[2]{#1 \, \constr{\langle} \, #2 \, \constr{\rangle} \, . \,}
\newcommand{\recv}[1]{#1 \, \constr{\mathbb{()}} \,}
\newcommand{\recvp}[2]{#1 \, \constr{(} \, #2 \, \constr{)} \, . \, }
\newcommand{\suc}{\constr{\scriptstyle 1+}}
\newcommand{\unit}{\constr{\mathbbm{1}}}
\newcommand{\base}[1]{\constr{B[} \, #1 \, \constr{]}}
\newcommand{\channel}[2]{\constr{C[} \, #1 \, \constr{,} \, #2 \, \constr{]}}
\newcommand{\comma}{\, \constr{,} \,}

@@

% Types
\newcommand{\Set}{\type{SET}}
\newcommand{\reduce}[1]{\, \type{\longrightarrow}_{#1} \,}
\newcommand{\types}[4]{#1 \, \type{;} \, #2 \, \type{\vdash} \, #3 \, \type{\triangleright} \, #4}
\newcommand{\contains}[6]{#1 \, \type{;} \, #2 \, \type{\ni}_{#3} \, #4 \, \type{;} \, #5 \, \type{\triangleright} \, #6}
\newcommand{\containsusage}[4]{#1 \, \type{\ni}_{#2} \, #3 \, \type{\triangleright} \, #4}
\newcommand{\Var}{\type{VAR}}
\newcommand{\Process}{\type{PROCESS}}
\newcommand{\Unused}{\type{UNUSED}}
\newcommand{\PreCtx}{\type{PRECTX}}
\newcommand{\Ctx}{\type{CTX}}
\newcommand{\Type}{\type{TYPE}\,}
\newcommand{\Idx}{\type{IDX}\,}
\newcommand{\Idxs}{\type{IDXS}}
\newcommand{\Usage}{\type{USAGE}}
\newcommand{\N}{\type{\mathbb{N}}}
\newcommand{\Channel}{\type{CHANNEL}}
\newcommand{\Rec}{\type{REC}}
\newcommand{\Algebra}{\type{ALGEBRA}}
\newcommand{\eq}[1]{\, \type{\simeq}_{#1} \,}
\newcommand{\eqeq}{\, \type{\cong} \,}

@@

\autoref{fig:notation} illustrates the notation used in this paper.
Data type definitions ($\type{\N}$) use double lines and index-free synonyms (\textsc{Nat}) as rule names (for ease of reference).
We otherwise use constructor names ($\constr{0}$ and $\suc$) to name our typing rules.
Universe levels and universe polymorphism are omitted for brevity --- all our types are of type $\Set$.
Implicit arguments are mentioned by type definitions but omitted by constructors.

@@

@@

      \datatype
      {n : \N}
      {\Process_n : \Set}
      \; \textsc{Process}
    \end{mathpar}
\end{figure}

\begin{figure}[h]
    \begin{equation*}
      \begin{split}
        P,Q ::=      & \; \PO                \\ 
                    |& \; (\new{}x) \; P       \\ 
                    |& \; \comp{P}{Q}        \\ 
                    |& \; \recvp{x}{y} P \\ 
                    |& \; \sendp{x}{y} P 
      \end{split}
    \end{equation*}
    \caption{Ill-scoped grammar with names.}
    \label{fig:syntax-names}
\end{figure}

\begin{figure}[h]
    \begin{equation*}
      \begin{aligned}
        \Process_n ::=& \; \PO_n               &&\text{(inaction)}   \\ 
        |& \; \new{} \Process_{\suc n}         &&\text{(restriction)}          \\ 
        |& \; \comp{\Process_n}{\Process_n}    &&\text{(parallel)} \\ 
        |& \; \recv{\Var_n}{}\Process_{\suc n} &&\text{(input)}                \\ 
        |& \; \send{\Var_n}{\Var_n}\Process_n  &&\text{(output)}                  
      \end{aligned}
    \end{equation*}
    \caption{Well-scoped grammar with type-level de Bruijn indices.}
    \label{fig:syntax}
\end{figure}

Let $x, y,\ldots$ range over \emph{variables} (including channel names) and $P, Q,\ldots$ over processes.
The syntax of the \picalc{} \cite{Sangio01} is given by the grammar in \autoref{fig:syntax-names}.
Process $\PO$ denotes the terminated process, where no operations can further occur.
Process $(\new{}x)\; P$ creates a new channel $x$ bound with scope $P$.
Process $\comp{P}{Q}$ is the parallel composition of processes $P$ and $Q$.
Processes $\recvp{x}{y} P$ and $\sendp{x}{y} P$ denote respectively the input and output processes of a variable $y$ over a channel $x$, with continuation $P$.

Scope restriction $(\new{}x) \; P$ and input $\recvp{x}{y} \; P$ are \emph{binders}, they are the only constructs which introduce bound names --- $x$ and $y$ in $P$, respectively.
In order to mechanise the \picalc{} syntax in Agda, we need to deal with bound names in continuation processes.
Names are cumbersome to mechanise: inserting a new variable into a context means proving that its name differs from all other variable names in context.
Instead, we use de Bruijn indices \cite{deBruijn1972}, where a natural number $n$ (aka \emph{index}) is used to refer to the variable introduced $n$ binders ago --- hence binders no longer introduce names.
\begin{example}
  Example process using names (top) and de Bruijn indices (bottom).
  \small
  \begin{equation*}
    \begin{aligned}
      & (\new{}x ) && (\comp {\recvp{x}{y} && \sendp{y}{a} && \PO} {(\new{} y) && (\sendp{x}{y} && \recvp{y}{z} && \PO)}) \\
      & \new{} && (\comp {\recv{0} && \send{0}{a} && \PO} {\new{} && (\send{1}{0} && \recv{0} && \PO)}) \\
    \end{aligned}
  \end{equation*}
\end{example}
That is, terms at different \emph{depths} use different indices to refer to the same binding.

A variable reference occurring under $n$ binders can refer to $n$ distinct variables.
References outside of that range are meaningless.
It is useful to rule out these ill-scoped nonsensical terms syntactically.
In \autoref{fig:syntax}, we do so by introducing the indexed family of types \cite{Dybjer1994} $\Var_n$: for all naturals $n$, the type $\Var_n$ has $n$ distinct elements.
We index processes according to their \emph{depth}: for all naturals $n$, a process of type $\Process_n$ contains variables that can refer to $n$ distinct elements.
Every time we go under a binder, we increase the index of the continuation process, allowing the variable references within to refer to one more thing.

@@

The operational semantics of the \picalc{} is defined as a reduction relation \textsc{Reduces} $\reduce{c}$ indexed by the channel $c$ on which communication occurs (\autoref{fig:reduction}).
We keep track of channel $c$ so we can state subject reduction (\autoref{thm:subject-reduction}).
We distinguish between channels that are created inside of the process ($\constr{internal}$) and channels that are created outside ($\constr{external}\ i$, where $i$ is the index of the channel variable).

@@

  \inferrule
  {i \; j : \Var_n \\ P : \Process_{\suc n} \\ Q : \Process_n}
  {\constr{comm} : \comp{\recv{i}P}{\send{i}{j}{Q}} \reduce{\constr{external} \; i} \comp{\func{lower}_{\constr{0}} \; (\subst{P}{\suc j}{\constr{0}}) \; uP'}{Q}}

@@

    \inferrule
        {\stacked{
            chan_i : \contains{\gamma \hspace{1.2em}}{\Gamma \hspace{1.3em}}{i}{\channel{t}{x}}{\li}{\Xi} \\\\
            cont \hspace{0.6em} : \types{\gamma \comma t}{\Xi \comma x}{P \hspace{4.6em}}{\Theta \comma \lz}}}
        {\constr{recv} \; chan_i \; cont : \types{\gamma}{\Gamma}{\recv{i}{P}}{\Theta}}

@@

  \begin{lstlisting}[label=lst:derivation,mathescape,numbers=left,caption=Typing a process that sends part of one channel over another.]
  $\func{p}$ : $\Process_{\suc \constr{zero}}$
  $\func{p}$ = $\new{}$
       $(\comp{\recv{\constr{0}} (\recv{\constr{0}} \PO)}$
       ${\new{} (\send{\suc \; \constr{0}} {\constr{0}} (\send{\constr{0}}{\suc \; \suc \; \constr{0}}) \; \PO)})$

>>>>

% Constructors
\newcommand{\PO}{\constr{\mathbb{0}}}
\newcommand{\comp}[2]{#1 \, \constr{\parallel} \, #2}
\newcommand{\new}{\constr{\boldsymbol{\nu}} \,}
\newcommand{\send}[2]{#1 \, \constr{\langle} \, #2 \,\constr{\rangle} \,}
\newcommand{\recv}[2]{#1 \, \constr{\mathbb{(}} \, #2 \, \constr{\mathbb{)}} \,}
\newcommand{\suc}{\constr{\scriptstyle 1+}}
\newcommand{\unit}{\constr{\mathbbm{1}}}
\newcommand{\base}[1]{\constr{B[} \, #1 \, \constr{]}}
\newcommand{\channel}[2]{\constr{C[} \, #1 \, \constr{,} \, #2 \, \constr{]}}
\newcommand{\comma}{\, \constr{,} \,}

@@

% Types
\newcommand{\Set}{\type{SET}}
\newcommand{\reduce}[1]{\, \type{\longrightarrow}_{#1} \,}
\newcommand{\types}[4]{#1 \, \type{;} \, #2 \, \type{\vdash} \, #3 \, \type{\triangleright} \, #4}
\newcommand{\contains}[6]{#1 \, \type{;} \, #2 \, \type{\ni}_{#3} \, #4 \, \type{;} \, #5 \, \type{\triangleright} \, #6}
\newcommand{\containsusage}[4]{#1 \, \type{\ni}_{#2} \, #3 \, \type{\triangleright} \, #4}
\newcommand{\Var}{\type{VAR}}
\newcommand{\Process}{\type{PROCESS}}
\newcommand{\Raw}{\type{RAW}}
\newcommand{\Name}{\type{NAME}}
\newcommand{\Names}{\type{NAMES}}
\newcommand{\Fresh}{\type{FRESH}}
\newcommand{\WellScoped}{\type{WELLSCOPED}}
\newcommand{\Unused}{\type{UNUSED}}
\newcommand{\PreCtx}{\type{PRECTX}}
\newcommand{\Ctx}{\type{CTX}}
\newcommand{\Type}{\type{TYPE}\,}
\newcommand{\Idx}{\type{IDX}\,}
\newcommand{\Idxs}{\type{IDXS}}
\newcommand{\Usage}{\type{USAGE}}
\newcommand{\N}{\type{\mathbb{N}}}
\newcommand{\Channel}{\type{CHANNEL}}
\newcommand{\Rec}{\type{REC}}
\newcommand{\Algebra}{\type{ALGEBRA}}
\newcommand{\eq}[1]{\, \type{\simeq}_{#1} \,}
\newcommand{\eqeq}{\, \type{\cong} \,}

@@

\autoref{fig:notation} illustrates the notation used in this paper.
Data type definitions ($\type{\N}$) use double lines and index-free synonyms (\textsc{Nat}) as rule names (for ease of reference).
Constructors ($\constr{0}$ and $\suc$) are given as inference rules, although we occasionally use a BNF style definition of types for brevity.
We maintain a close correspondence between the definitions presented in this paper and our mechanised definitions in Agda --- where premises become argument types and conclusions return types.
Universe levels and universe polymorphism are omitted for brevity --- all our types are of type $\Set$.
Implicit arguments are mentioned by type definitions but omitted by constructors.

@@

\begin{figure}[h]
  \begin{mathpar}
    \datatype
    { }
    {\Raw : \Set}
    \; \textsc{Raw}
  \end{mathpar}

  \begin{equation*}
    \begin{aligned}
      \Raw ::= &\; \PO              &&\text{(inaction)}    \\ 
      |& \; (\new{\Name}) \; \Raw         &&\text{(restriction)} \\ 
      |& \; \comp{\Raw}{\Raw}       &&\text{(parallel)}    \\ 
      |& \; \recv{\Name}{\Name}\Raw &&\text{(input)}       \\ 
      |& \; \send{\Name}{\Name}\Raw &&\text{(output)}      \\
    \end{aligned}
  \end{equation*}
  \caption{Grammar with names.}
  \label{fig:syntax-names}
\end{figure}

@@

    \datatype
    {n : \N}
    {\Process_n : \Set}
    \; \textsc{Process}
  \end{mathpar}

  \begin{equation*}
    \begin{aligned}
      \Process_n ::=& \; \PO_n               &&\text{(inaction)}   \\ 
      |& \; \new{} \Process_{\suc n}         &&\text{(restriction)}          \\ 
      |& \; \comp{\Process_n}{\Process_n}    &&\text{(parallel)} \\ 
      |& \; \recv{\Var_n}{}\Process_{\suc n} &&\text{(input)}                \\ 
      |& \; \send{\Var_n}{\Var_n}\Process_n  &&\text{(output)}                  
    \end{aligned}
  \end{equation*}
  \caption{Well-scoped grammar with type-level de Bruijn indices.}
  \label{fig:syntax}
\end{figure}

The syntax of the \picalc{} \cite{Sangio01} is given by the grammar in \autoref{fig:syntax-names}.
Process $\PO$ denotes the terminated process, where no operations can further occur.
Process $(\new{}x)\; P$ creates a new channel $x$ bound with scope $P$.
Process $\comp{P}{Q}$ is the parallel composition of processes $P$ and $Q$.
Processes $\recv{x}{y} P$ and $\send{x}{y} P$ denote respectively the input and output processes of a variable $y$ over a channel $x$, with continuation $P$.

Scope restriction $(\new{}x) \; P$ and input $\recv{x}{y} \; P$ are \emph{binders}, they are the only constructs which introduce bound names --- $x$ and $y$ in $P$, respectively.
In order to mechanise the \picalc{} syntax in Agda, we need to deal with bound names in continuation processes.
Names are cumbersome to mechanise: they offer no guarantee of being well-scoped, one has to deal with alpha-conversion, and inserting new variables into context entails proving that their names differ from all other names in context.
Instead, we use de Bruijn indices \cite{deBruijn1972}, where a natural number $n$ (aka \emph{index}) is used to refer to the variable introduced $n$ binders ago.
That is, terms at different \emph{depths} use different indices to refer to the same binding, and binders no longer introduce names.
\begin{example}
  Example process using names (top) and de Bruijn indices (bottom).
  \small
  \begin{equation*}
    \begin{aligned}
      & (\new{}x ) && (\comp {\recv{x}{y} && \send{y}{a} && \PO} {(\new{} y) && (\send{x}{y} && \recv{y}{z} && \PO)}) \\
      & \new{} && (\comp {\recv{0}{} && \send{0}{a} && \PO} {\new{} && (\send{1}{0} && \recv{0}{} && \PO)}) \\
    \end{aligned}
  \end{equation*}
\end{example}

A variable reference occurring under $n$ binders can refer to $n$ distinct variables.
References outside of that range are meaningless.
It is useful to rule out these ill-scoped terms syntactically.
In \autoref{fig:syntax}, we do so by introducing the indexed family of types \cite{Dybjer1994} $\Var_n$: for all naturals $n$, the type $\Var_n$ has $n$ distinct elements.
We index processes according to their \emph{depth}: for all naturals $n$, a process of type $\Process_n$ contains variables that can refer to $n$ distinct elements.
Every time we go under a binder, we increase the index of the continuation process, allowing the variable references within to refer to one more thing.

\subsection{From names to de Bruijn indices and back}

De Bruijn indices are easier to manipulate, but as humans we rather deal with names.
To that end, we provide the user with named processes, while using de Bruijn indexed processes internally.
We do so by defining translation functions in both directions and proving that they are inverses of each other up to $\alpha$-conversion.

\paragraph{From names to de Bruijn indices}
We keep the original binder names around as name hints when we translate into processes with de Bruijn names --- these names will serve as hints when we translate back.
The translation function $\func{fromRaw}$ works recursively, keeping a context $ctx : \Names_n$ that maps between the first $n$ indices and their names, and substituting named references within the process with their corresponding de Bruijn index.
We demand that the original process is well-scoped: that all its free variable names appear in $ctx$ --- this is decidable and we therefore automate it.
\begin{equation*}
  \begin{aligned}
    &\func{fromRaw} && : (ctx : \Names_n) \; (P : \Raw) \\
    &               && \to \WellScoped \; ctx \; P \to \Process_n \\
  \end{aligned}
\end{equation*}

\paragraph{From de Bruijn indices to names}
The translation function $\func{toRaw}$ works recursively, keeping a context $ctx : \Names_n$ that maps between $n$ de Bruijn indices and names, and substituting indices within the process with their corresponding names.
As many widely-used languages do, we produce unique variable names but use the original variable names kept around throughout the de Bruijn representation as hints, to make the correspondence visually more straightforward.
For this, we use the naming scheme $<name-hint>\textasciicircum<n>$, where $<n>$ denotes that the name $<name-hint>$ has already been bound $n$ times before.
\begin{equation*}
  \begin{aligned}
    &\func{toRaw} && : (ctx : \Names_n) \to \Process_n \to \Raw
  \end{aligned}
\end{equation*}

\begin{nilemma}
  Translating from de Bruijn indices to names results in a well-scoped process.
\end{nilemma}

\begin{nilemma}
  Translating from de Bruijn indices to names results in a process with no shadowed variables.
\end{nilemma}

\begin{nilemma}
  Translating from de Bruijn indices to names and back results in the same process modulo variable name hints.
\end{nilemma}

\begin{nilemma}
  Translating from names to de Bruijn indices and back results in the same process modulo $\alpha$-conversion.
\end{nilemma}

\begin{proof}
  TODO
\end{proof}

@@

The operational semantics of the \picalc{} is defined as a reduction relation \textsc{Reduces} $\reduce{c}$ indexed by the channel $c$ on which communication occurs (\autoref{fig:reduction}).
We keep track of channel $c$ so we can state subject reduction (\autoref{thm:subject-reduction}).
We distinguish between channels that are created inside of the process ($\constr{internal}$) and channels that are created outside ($\constr{external}\ i$), where $i$ is the index of the channel variable).

@@

  \inferrule
  {i \; j : \Var_n \\ P : \Process_{\suc n} \\ Q : \Process_n}
  {\constr{comm} : \comp{\recv{i}{}P}{\send{i}{j}{Q}} \reduce{\constr{external} \; i} \comp{\func{lower}_{\constr{0}} \; (\subst{P}{\suc j}{\constr{0}}) \; uP'}{Q}}

@@

    \inferrule
        {\stacked{
            chan_i : \contains{\gamma \hspace{1.2em}}{\Gamma \hspace{1.3em}}{i}{\channel{t}{x}}{\li}{\Xi} \\\\
            cont \hspace{0.6em} : \types{\gamma \comma t}{\Xi \comma x}{P \hspace{4.6em}}{\Theta \comma \lz}}}
        {\constr{recv} \; chan_i \; cont : \types{\gamma}{\Gamma}{\recv{i}{}P}{\Theta}}

@@

  \begin{lstlisting}[label=lst:derivation,mathescape,numbers=left,caption=Typing a process that sends part of one channel over another.]
  $\func{p}$ : $\Process_{\suc \constr{zero}}$
  $\func{p}$ = $\new{}$
       $(\comp{\recv{\constr{0}}{} (\recv{\constr{0}}{} \PO)}$
       ${\new{} (\send{\suc \; \constr{0}} {\constr{0}} (\send{\constr{0}}{\suc \; \suc \; \constr{0}}) \; \PO)})$
